ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2012-01-05 papers/IJCSE/programming-model.tex
one more edit from Bob

The High Performance Fortran (HPF) extension of Fortran was proposed
to add features to the language that would enhance the ability of
compilers to emit fast parallel code for distributed and shared memory
parallel computers\citep{koelbel94hpf}.  One of the notable additions
to the language in HPF was syntax to specify the distribution of data
structures amongst a set of parallel processors.  HPF also introduced
an alternative looping construct to the traditional DO-loop called
{\tt FORALL} that was better suited for parallel compilation.  An
additional keyword, {\tt INDEPENDENT}, was added to allow the
programmer to indicate when the loop contained no loop-order
dependencies that allowed for parallel execution.  These constructs
are similar to {\tt DO CONCURRENT}, an addition to Fortran in 2008.

@@

Unfortunately, this had two side effects that have limited broad
acceptance of the whole-array programming model in Fortran.  First,
programmers must translate their algorithms into a set of global
operations.  Finite difference stencils and similar computations are
traditionally defined in terms of offsets from some central index.
Shifting, while conceptually analogous, can be awkward to think about
for high dimensional stencils with many points.  Second, the semantics
of these operations are such that all elements of an array operation
are updated as if they were updated simultaneously.  In a program
where the programmer explicitly manages arrays and loops, double
buffering techniques and user managed temporaries are used to maintain
these semantics.  Limited attention to optimizing memory usage due to
this intermediate storage by compilers has led to these constructs
seeing little adoption by programmers.

@@

It should be noted that the function {\tt region\_cpy} is pure and
thus can be called from within a pure kernel procedure, and {\tt
  region\_ptr} is impure because it aliases the array parameter.
However as will be shown below, the usage of {\tt region\_ptr} is
constrained so that it does not introduce side effects in the
functions that call it.  These two functions are part of the language
recognized by the compiler and though {\tt region\_cpy} returns a copy
of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics
through other means within the pure constraint.  In addition to these
two new functions, ForOpenCL provides the compiler directive, {\tt
  \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine can be
transformed to an OpenCL kernel and that the subroutine is pure except
for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in
automated identification of specific kernels to be transformed to
OpenCL.  A directive-free implementation would require the
transformation tool be provided the set of kernels to work via a user
defined list.

@@

It is assumed that memory for all arrays reside on the device.  The
programmer must copy memory to and from the device.  In addition,
array size (neglecting ghost cell regions) must be multiples of the
global OpenCL kernel size.

Array variables within a kernel procedure (specified by the {\tt
  \!\$OFP PURE, KERNEL} directive) must be declared as contiguous.  A
kernel procedure may not call other procedures except for limited
intrinsic functions (primarily math), user-defined elemental
functions, and the {\tt region\_cpy} and {\tt region\_ptr} functions.
Future work will address non-contiguous arrays (such as those that
result from strided access) by mapping array strides to
gather/scatter-style memory accessors.

>>>>

The High Performance Fortran (HPF) extension of Fortran was proposed
to add features to the language that would enhance the ability of
compilers to emit fast parallel code for distributed and shared memory
parallel computers~\citep{koelbel94hpf}.  One of the notable additions
to the language in HPF was syntax to specify the distribution of data
structures amongst a set of parallel processors.  HPF also introduced
an alternative looping construct to the traditional DO-loop called
{\tt FORALL} that was better suited for parallel compilation.  An
additional keyword, {\tt INDEPENDENT}, was added to allow the
programmer to indicate when the loop contained no loop-order
dependencies that allowed for parallel execution.  These constructs
are similar to {\tt DO CONCURRENT}, an addition to Fortran in 2008.

@@

Unfortunately, this had two side effects that have limited broad
acceptance of the whole-array programming model in Fortran.  First,
programmers must translate their algorithms into a set of global
operations.  Finite difference stencils and similar computations are
traditionally defined in terms of offsets from some central index.
Shifting, while conceptually analogous, can be awkward to think about
for high dimensional stencils with many points.  Second, the semantics
of these operations are such that all elements of an array operation
are updated as if they were updated simultaneously.  In a program
where the programmer explicitly manages arrays and loops (rather than
the compiler), double buffering techniques and user managed
temporaries are used to maintain these semantics.  Limited attention
to optimizing memory usage due to this intermediate storage by
compilers has led to these constructs seeing little adoption by
programmers.

@@

It should be noted that the function {\tt region\_cpy} is pure and
thus can be called from within a pure kernel procedure, and {\tt
  region\_ptr} is impure because it aliases the array parameter.
However as will be shown below, the usage of {\tt region\_ptr} is
constrained so that it does not introduce side effects in the
functions that call it.  These two functions are part of the language
recognized by the compiler and though {\tt region\_cpy} returns a copy
of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics
through other means within the pure constraint.  In addition to these
two new functions, ForOpenCL provides the compiler directive, {\tt
  \!\$OFP PURE, CONCURRENT}, which specifies that a pure subroutine can be
transformed to an OpenCL kernel and that the subroutine is pure except
for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in
automated identification of specific kernels to be transformed to
OpenCL.  A directive-free implementation would require the
transformation tool be provided the set of kernels to work via a user
defined list.

@@

It is assumed that memory for all arrays reside on the device.  The
programmer must copy memory to and from the device.  In addition,
array size (neglecting ghost cell regions) must be multiples of the
global OpenCL kernel size.  The explicit memory copy and array sizes
are constraints of this current work and not of the possible future
enhancements.

Array variables within a kernel procedure (specified by the {\tt
  \!\$OFP PURE, CONCURRENT} directive) must be declared as contiguous.  A
kernel procedure may not call other procedures except for limited
intrinsic functions (primarily math), user-defined elemental
functions, and the {\tt region\_cpy} and {\tt region\_ptr} functions.
Future work will address non-contiguous arrays (such as those that
result from strided access) by mapping array strides to
gather/scatter-style memory accessors.

### 2012-01-05 papers/IJCSE/shallow-water.tex

The numerical method is a two-step Lax-Wendroff scheme. The method
has some numerical oscillations with sharp gradients but is adequate
for simulating smooth shallow-water flows. In the following explanation,
$U$ is the conserved state variable at the center of the cell. This
state variable, $U$ $=\{h,hu,hv\}$ in the first term in the equations
below. $F$ is the flux quantity that crosses the boundary of the cell
and is subtracted from one cell and added to the other. The remaining
terms after the first term are the flux terms in the equations above
with one term for the flux in the x-direction and the next term for
the flux in the y-direction. The first step estimates the values a
half-step advanced in time and space on each face, using loops on
the faces.\begin{eqnarray*}
U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)\end{eqnarray*}

The second step uses the estimated values from step 1 to compute the
values at the next time step in a dimensionally unsplit loop.\[
U_{i,j}^{n+1}=U_{i,j}^{n}-\frac{\triangle t}{\triangle x}(F_{i+\frac{1}{2},j}^{n+\frac{1}{2}}-F_{i-\frac{1}{2},j}^{n+\frac{1}{2}})-\frac{\triangle t}{\triangle y}(F_{i,j+\frac{1}{2}}^{n+\frac{1}{2}}-F_{i,j-\frac{1}{2}}^{n+\frac{1}{2}})\]

>>>>

The numerical method is a two-step Lax-Wendroff scheme. The method
has some numerical oscillations with sharp gradients but is adequate
for simulating smooth shallow-water flows. In the following explanation,
$U$ is the conserved state variable at the center of the cell. This
state variable, $U$ $=\{h,hu,hv\}$ in the first term in the equations
below. $F$ is the flux quantity that crosses the boundary of the cell
and is subtracted from one cell and added to the other. The remaining
terms after the first term are the flux terms in the equations above
with one term for the flux in the x-direction and the next term for
the flux in the y-direction. The first step estimates the values a
half-step advanced in time and space on each face, using loops on
the faces.

\begin{eqnarray*}
U_{i+\frac{1}{2},j}^{n+\frac{1}{2}} & = & (U_{i+1,j}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle x}\left(F_{i+1,j}^{n}-F_{i,j}^{n}\right)\\
U_{i,j+\frac{1}{2}}^{n+\frac{1}{2}} & = & (U_{i,j+1}^{n}+U_{i,j}^{n})/2+\frac{\triangle t}{2\triangle y}\left(F_{i,j+1}^{n}-F_{i,j}^{n}\right)
\end{eqnarray*}

The second step uses the estimated values from step 1 to compute the
values at the next time step in a dimensionally unsplit loop.

\begin{eqnarray*}
U_{i,j}^{n+1} = U_{i,j}^{n} &-&\frac{\triangle t}{\triangle
  x}(F_{i+\frac{1}{2},j}^{n+\frac{1}{2}}-F_{i-\frac{1}{2},j}^{n+\frac{1}{2}})\\
&-& \frac{\triangle
  t}{\triangle
  y}(F_{i,j+\frac{1}{2}}^{n+\frac{1}{2}}-F_{i,j-\frac{1}{2}}^{n+\frac{1}{2}})
\end{eqnarray*}

### 2012-01-05 papers/IJCSE/source-to-source.tex
edits

OpenCL~\cite{opencl08} is an open-language standard for developing applications
targeted for GPUs, as well as for multi-threaded applications targeted for
multi-core CPUs.  The kernels are run by calling a C runtime library from the
OpenCL host (normally the CPU).  Efforts to standardize a C++ runtime are
underway and Fortran interfaces to the C runtime are distributed in the ForOpenCL
library.

@@

\vskip .4in

@@

>>>>

OpenCL~\citep{opencl08} is an open-language standard for developing
applications targeted for GPUs, as well as for multi-threaded
applications targeted for multi-core CPUs.  The kernels are run by
calling a C runtime library from the OpenCL host (normally the CPU).
Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are distributed in the ForOpenCL library.

@@

@@

\input{rose-transform}
